---
title: "Whiskerbook Data Manipulation"
objectives:
- "Perform data wrangling and manipulation"
- "Sort only high quality data"
- "Divide data into left only and right only"
questions: 
- "How to manipulate data for descriptive information"
teaching: 60
exercises: 30
source: Rmd
keypoints:
- "Subset left and right only individuals"
- "Subset individuals that have encounters that include both left and right sides"
- "Subset dataframes that include left only and both sides, and right and both sides"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-libraries, echo = FALSE, results='hide', warning = FALSE, message = FALSE}
library(dplyr)
edf_tdf<-read.csv("edf_tdf.csv")
```

Let's start by sorting the juvenilles out of our dataframe.
```{r}
#subset the dataframe for records containing juvenilles
edf_tdf_J<-edf_tdf[which(edf_tdf$Juvenilles=="J"),]

#subset the dataframe to identify unique individuals
edf_tdf_J<-unique(edf_tdf_J[,c("Marked.Individual","date_Time")])

#pull the indiviudal ID out
edf_tdf_J_ID<-unique(as.character(edf_tdf_J$Marked.Individual))

```

Lets remove the Juvenilles from our dataframe, since we won't need them for our analysis our our descriptions. 
```{r}
#Remove juvenilles from the dataframe
edf_tdf<-edf_tdf[-which(edf_tdf$Juvenilles=="J"),]
```

Subset the data by distinct fields, since the encounter data can be inflated because the Whiskerbook program can include several unique encounters for the same individual.
```{r}
#get unique records by several fields using distinct
edf_tdf_d<-edf_tdf %>% distinct(Marked.Individual, date_Time, Side, Location.ID, .keep_all = TRUE)
```

Next we can create some dataframes of high quality and low quality encounters.
```{r}
edf_tdf_high<-edf_tdf_d[which(edf_tdf_d$Quality=="H"),]
edf_tdf_low<-edf_tdf_d[which(edf_tdf_d$Quality %in% c("L","M")),]
```

Now we will do some basic descriptions of the left and right hand sides and determine which individuals have left sides and right sides, or only one or the other. 
```{r}
ID_left_right<-edf_tdf_high%>%
  group_by(Marked.Individual, Side)%>%
  count()
```
  
Let's subset the data that are annotated with either left and right.
```{r}
ID_left_right_sub<-ID_left_right[which(ID_left_right$Side %in% c("L", "R")),]
```

Lets see how many individuals have only the left side or right side by counting the records grouped by ID. 
```{r}
ID_left_right_sub_final<-ID_left_right_sub%>%
  group_by(Marked.Individual)%>%
  count()
```

From this,we can find the records which are left only or right only individuals. 
```{r}
OneSided<-ID_left_right_sub_final[which(ID_left_right_sub_final$n==1),]
TwoSided<-ID_left_right_sub_final[which(ID_left_right_sub_final$n==2),]
```

Now we can find which individuals have one side or both sides.
```{r}
ID_oneOnly<-ID_left_right_sub[which(ID_left_right_sub$Marked.Individual %in% OneSided$Marked.Individual),]
ID_twoSided<-ID_left_right_sub[which(ID_left_right_sub$Marked.Individual %in% TwoSided$Marked.Individual),]
```

From this we can find the individuals that are one sided and right only or left only.
```{r}
ID_oneOnly_right<-ID_oneOnly[ID_oneOnly$Side=="R",]
ID_oneOnly_left<-ID_oneOnly[ID_oneOnly$Side=="L",]
```

Now we can subset the dataframe to remove the records that are right only to create the left only dataframe. Similarly, when we subtract out the left only records, then we have a right only dataframe.
```{r}
edf_left<-edf_tdf_high[-which(edf_tdf_high$Marked.Individual %in% ID_oneOnly_right$Marked.Individual),]
edf_right<-edf_tdf_high[-which(edf_tdf_high$Marked.Individual %in% ID_oneOnly_left$Marked.Individual),]
```




> ## Challenge: Camera Operability Matrix

> 
> Answer the following questions:
> 
> 1. How many individuals were back only?
> >ID_back_final<-ID_left_right%>%
  group_by(Side)%>%
  count()
  
>>From this result, we can see that 9 detections were back, and 3 were tail only.


```

> {: .solution}
{: .challenge}

{% include links.md %}


