---
title: "Spatial-Capture Recapture Modeling"
objectives:
- "Perform single session spatial capture recapture modeling tasks"
- "Read outputs for density, abundance, detectability and sigma"

questions: 
- "How to setup and run oSCR models?"
- "How to interpret the model outputs?"

teaching: 60
exercises: 30
source: Rmd
---


```{r setup, echo=FALSE}
source("../bin/chunk-options.R")
source("../setup.R")
knitr_fig_path("06-")
```

Use the oSCR.fit function with no covariates, use the scrFrame and alltraps_df that we generated earlier.

Then use predict.oSCR onto the same data to get our predictions. 

Note that this will take around 5 minutes to run.

```{r}
snowLeopard.1<- oSCR.fit(list(D ~ 1, p0 ~ 1, sig ~ 1), scrFrame, list(alltraps_df))
pred<-predict.oSCR(snowLeopard.1, scrFrame,list(alltraps_df), override.trim =TRUE )
```


We can plot the estimates for density across the study area to see how it looks
```{r}
library(viridis)
myCol = viridis(7)
RasterValues_1<-as.matrix(pred$r[[1]])
MaxRaS<-max(RasterValues_1, na.rm=TRUE)
MinRaS<-min(RasterValues_1,na.rm=TRUE)

plot(pred$r[[1]], col=myCol,
     main="Realized density",
     xlab = "UTM Westing Coordinate (m)", 
     ylab = "UTM Northing Coordinate (m)")
points(tdf2[,3:4], pch=20)
```


Backtransforming the estimates to be in the 100km2 units for density that we want using ht emu
```{r}
pred.df.dens <- data.frame(Session = factor(1))
#make predictions on the real scale
(pred.dens <- get.real(snowLeopard.1, type = "dens", newdata = pred.df.dens, d.factor = multiplicationfactor))
```

Get the abundance, detection, and sigma parameters
```{r}
(total.abundance <- get.real(snowLeopard.1, type = "dens", newdata = pred.df.dens, d.factor=nrow(snowLeopard.1$ssDF[[1]])))
```

```{r}
(pred.det <- get.real(snowLeopard.1, type = "det", newdata = pred.df.dens))
```

```{r}
(pred.sig <- get.real(snowLeopard.1, type = "sig", newdata = pred.df.dens))
```